---
title: "AlberdiLab | Marcos et al. 2025"
subtitle: "Title to be added"
author:
  - Sofia Marcos, Iñaki Odriozola, Ostaizka Aizpurua, Raphael Eisenhofer, Sarah Siu Tze Mak, Garazi Martin, Varsha Kale, Germana Baldi, Robert D Finn, Joan Tarradas, Andone Estonba, M Thomas P Gilbert, Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/chicken_genome_reduced_bacteria
description: |
  Data analysis code developed to study early-life caecal microbiome development and its association with chicken body weight 
link-citations: yes
github-repo: alberdilab/chicken_genome_reduced_bacteria
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Chapter 1 - Introduction

This WebBook contains the code written to analyse data from the study of the functional profile of metagenome-assembled genomes and temporal trends of derived microbial communities from intensively reared broiler chickens.

## 1.1 - Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/chicken_genome_reduced_bacteria.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning = FALSE, comments = "", message = FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(rmarkdown)
library(janitor)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)
library(tidytree)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)
library(sjPlot)
library(colorspace)
library(vioplot)

# For statistics
library(spaa)
library(vegan)
library(Rtsne)
library(microbiome)
library(geiger)
library(hilldiv2)
library(distillR)
library(broom.mixed)
library(Hmsc)
library(MuMIn)
library(corrplot)
library(lme4)
library(nlme)
library(boot)
```
