# Chapter 2  - Metagenomic data preparation

## 2.1 - Load data

Load the original data files outputted by the bioinformatic pipeline.

### Chicken sample metadata

```{r load_metadata, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
chicken_metadata <- 
  read_tsv("data/metadata.tsv") %>%
  mutate(sampling_time = factor(sampling_time, levels = c('7',
                                                          '21',
                                                          '35')))
```

### Taxonomy of metagenome-assembled genomes 

```{r load_taxonomy, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
genome_taxonomy <- 
  read_tsv("data/taxonomy.tsv") %>%
  rename(genome = 1) %>%
  mutate(phylum = case_when(
        phylum == "Actinobacteriota" ~ "Actinomycetota",
        phylum == "Firmicutes" ~ "Bacillota",
        phylum == "Firmicutes_A" ~ "Bacillota_A",
        phylum == "Firmicutes_B" ~ "Bacillota_B",
        phylum == "Firmicutes_C" ~ "Bacillota_C",
        phylum == "Cyanobacteria" ~ "Cyanobacteriota",
        phylum == "Proteobacteria" ~ "Pseudomonadota",
        TRUE ~ phylum)) %>% 
  filter(!domain == "Archaea")
```

### Genome statistics

```{r load_stats, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
genome_stats <- 
  read_tsv("data/stats.tsv") %>%
  rename(genome = 1) %>%
  mutate(correction_factor = median(mag_length) / mag_length) %>% 
  filter(genome %in% genome_taxonomy$genome) %>% 
  arrange(match(genome, genome_taxonomy$genome))
```

### Phylogenetic tree

```{r load_tree, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
genome_tree <- read_tree("data/tree.nwk")
```

### Raw gene annotations

```{r load_annotations, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
genome_annotations <- read_tsv("data/gene_annotations.tsv.xz")
```

### DRAM functional annotations

```{r load_dram_table, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
genome_kegg_paths <-
  read_tsv("data/dram.tsv") %>%
  rename(genome = 1) %>% 
  filter(genome %in% genome_taxonomy$genome) %>% 
  select(1:400)
```

### Metagenomic read counts

```{r load_read_counts, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
read_counts <- 
  read_tsv("data/mag_counts.tsv") %>% 
  rename(genome = 1) %>% 
  filter(genome %in% genome_taxonomy$genome) 
```


## 2.2 - Distilling Genome Inferred Functional Traits (GIFTs)

### Distilling DRAM annotations

```{r calculate_traits, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
gifts_raw <- distill(genome_annotations, GIFT_db2, genomecol = 2, annotcol = c(9,10,19), verbosity = F)

# Load ENA to genome_id table
ena_to_mag_id <- read_tsv("data/ena_to_mag_id.tsv")

gifts_matrix <-
  gifts_raw %>%
  data.frame() %>%
  rownames_to_column(var = 'mag_name') %>%
  left_join(ena_to_mag_id %>%
              select(mag_name, mag_id),
            by = 'mag_name') %>%
  select(-mag_name) %>%
  relocate(mag_id) %>%
  filter(mag_id %in% genome_taxonomy$genome) %>% 
  column_to_rownames('mag_id') %>%
  as.matrix()

genome_completeness <-
  genome_stats %>%
  select(genome, completeness_score) %>%
  as.matrix()
```

### Define completeness correction command

This correction is copied from https://github.com/anttonalberdi/DAMMA/blob/main/R/damma_correction.R, which is based on a previous publication https://www.nature.com/articles/s43705-023-00221-z.

```{r define_corretion, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
gifts_correction <- 
  function(MCI_table, genome_completeness, stats = TRUE) {
  # Sort Genomes and test matching
  MCI_table <- MCI_table[order(rownames(MCI_table)),]
  genome_completeness <- genome_completeness[order(genome_completeness[,1]),]
  if (all(as.character(rownames(MCI_table)) !=
          as.character(genome_completeness[,1])))
    stop("Pathway table is missing")
  
  # Get completeness values
  Genome_completeness <- as.numeric(genome_completeness[,2])
  
  # Create corrected fullness matrix
  MCI_table_corrected <- matrix(0, nrow = nrow(MCI_table), ncol = ncol(MCI_table))
  colnames(MCI_table_corrected) = colnames(MCI_table)
  rownames(MCI_table_corrected) = rownames(MCI_table)
  
  # Iterate modelling and correction for each function
  suppressWarnings(
    for (i in 1:ncol(MCI_table)) {
      Model <- glm(MCI_table[,i]~Genome_completeness,family = "binomial")
      slope_coef <- coef(Model)[2]
      if (slope_coef > 0) {
        for (j in 1:nrow(MCI_table)) {
          
          # Model prediction of fullness if completeness was 100%
          pred_100 <- round(predict(Model,
                                    newdata = data.frame(
                                      Genome_completeness = 100),
                                    type = "response"),1)
          
          # Model prediction of fullness for actual completeness of the focal MAG
          pred_focal <- round(predict(Model,
                                      newdata = data.frame(
                                        Genome_completeness = Genome_completeness[j]),
                                      type = "response"), 1)
          
          # The expected change in function fullness if focal MAG was 100% complete
          pred_diff <- pred_100 - pred_focal
          MCI_table_corrected[j,i] = MCI_table[j,i] + pred_diff
        }
      } else if (slope_coef <= 0) {
        MCI_table_corrected[,i] <- MCI_table[,i]
      }
    }
  )
  
  # If corrected fullness >1, convert it to 1.
  MCI_table_corrected[MCI_table_corrected > 1] <- 1
  
  # Outout overall correction statistics on screen
  total <- nrow(MCI_table)*ncol(MCI_table)
  changes <- c(MCI_table == MCI_table_corrected)
  changes <- length(changes[!(changes)])
  percentage <- round(changes / total * 100,1)
  cat(paste0(changes," out of ",total," (",percentage,"%) fullness values were corrected\n"))
  if (stats == TRUE) {
    
    # Outout overall correction statistics on screen
    total <- ncol(MCI_table_corrected)
    for (r in rownames(MCI_table_corrected)) {
      completeness <- round(as.numeric(genome_completeness[genome_completeness[,1] == r,2]),1)
      changes <- MCI_table[r,] == MCI_table_corrected[r,]
      changes <- length(changes[!(changes)])
      percentage <- round(changes / total * 100,1)
      cat(paste0("\t",r," (",completeness,"%): ",changes,
                 "/",total," (",percentage,"%) fullness values were corrected\n"))
    }
  }
  
  # Output corrected table
  return(MCI_table_corrected)
}
```

### Perform completeness correction for GIFTs

```{r perform_completeness_corretion, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
gifts_corrected <- gifts_correction(gifts_matrix, genome_completeness)

# Aggregate into compound level
gifts_elements <- to.elements(gifts_corrected, GIFT_db2)

# Aggregate into function level
gifts_functions <- to.functions(gifts_elements, GIFT_db2)

# Aggregate into overall domain level
gifts_domains <- to.domains(gifts_functions, GIFT_db2)
```


## 2.3 - Create working objects

Transform the original data files into working objects for downstream analyses.

### Transform reads into genome counts

```{r weighted_counts, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mag_weighted <-
  read_counts %>% 
  column_to_rownames(var = 'genome') %>% 
  mutate(across(where(is.numeric), ~ ./ genome_stats$correction_factor)) %>% 
  t() %>% 
  as.data.frame()
```

### Calculate relative abundances

```{r relative_abundance_tables, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# Relative abundances
total <-
  decostand(mag_weighted, 'total') %>%
  t() %>%
  as.data.frame()  

# Square root of relative abundances
hel <-
  decostand(mag_weighted, 'hellinger') %>%
  t() %>%
  as.data.frame() #¿? I kept both hel and total tables, but shall we just use one of them for all the analyses?

# Transpose matrix
mag_weighted <-
  mag_weighted %>% 
  t() %>% 
  as.data.frame()
```

### Prepare a phyloseq object

```{r phyloseq_objects, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
phylo_samples <- 
  chicken_metadata %>% 
  mutate(sampling_time = as.factor(sampling_time),
         trial_age = paste(trial, sampling_time, sep = "_"),
         trial_age = as.factor(trial_age)) %>% 
  column_to_rownames("animal_code") %>% 
  sample_data() # Convert to phyloseq sample_data object

phylo_genome <- otu_table(mag_weighted, taxa_are_rows = TRUE) 

phylo_taxonomy <- 
  genome_taxonomy %>%
  column_to_rownames("genome") %>% 
  as.matrix() %>% 
  tax_table() # Convert to phyloseq tax_table object

phylo_tree <- phy_tree(genome_tree) 

phylo_seq <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples, phylo_tree)
```

## 2.4 - Prepare color scheme

```{r get_mag_colors, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# as dataframe
taxonomy_colors <- read_tsv("data/taxonomy_colours.tsv") 

# as vector
phylum_colors <- 
  c(Halobacteriota = "#f2f2f2",
    Methanobacteriota = "#bfbfbf",
    Patescibacteria = "#dacce3",
    Campylobacterota = "#cdadca",
    Synergistota = "#c08eb3",
    Thermoplasmatota = "#777dae",
    Verrucomicrobiota = "#0066ae",
    Desulfobacterota = "#68b0dc",
    Pseudomonadota = "#60cfde",
    Bacteroidota = "#4dc87c",
    Cyanobacteriota = "#92e09f",
    Actinomycetota = "#c9e0af",
    Deferribacterota = "#dff77e",
    Bacillota = "#ffd366",
    Bacillota_A = "#fd8854",
    Bacillota_B = "#8b1222",
    Bacillota_C = "#5a0c37")

order_colors <- 
  c(Methanomicrobiales = "#f2f2f2",
    Methanobacteriales = "#bfbfbf",
    Saccharimonadales = "#dacce3",
    Campylobacterales = "#cdadca",
    Synergistales = "#c08eb3",
    Methanomassiliicoccales = "#777dae",
    Victivallales = "#0066ae",
    Opitutales = "#1c7ebc",
    Verrucomicrobiales = "#4a96cc",
    Desulfovibrionales = "#68b0dc",
    Enterobacterales = "#60cfde",
    RF32 = "#60dfd2",
    Burkholderiales = "#5cdfb5",
    'Rs-D84' = "#40df91",
    Bacteroidales = "#4dc87c",
    Flavobacteriales = "#88c88b",
    Gastranaerophilales = "#92e09f",
    Coriobacteriales = "#c9e0af",
    Actinomycetales = "#d8e093",
    Deferribacterales = "#dff77e",
    RF39 = "#ecf76d",
    Lactobacillales = "#feef68",
    'ML615J-28' = "#fde671",
    Erysipelotrichales = "#ffd366",
    Acholeplasmatales = "#fdc151",
    Bacillales = "#fc953d",
    RFN20 = "#fd8035",
    Oscillospirales = "#fd8854",
    TANB77 = "#f4814d",
    Christensenellales = "#c26340",
    Lachnospirales = "#c28c5c",
    Clostridiales = "#ce9360",
    Monoglobales_A = "#ce734f",
    Peptostreptococcales = "#c65631",
    Monoglobales = "#b93725",
    UBA1212 = "#b10f19",
    UBA4068 = "#8b1222",
    Selenomonadales = "#7a1a12",
    Acidaminococcales = "#64121f",
    Veillonellales = "#5a0c37")
```

```{r get_sampling_colors, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
sampling_day_colors <- c('#E69F00', '#CC6677', '#56B4E9')
```


## 2.5 - Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_mg_objects, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
save(
  # Chicken data
  chicken_metadata,
  
  # Bacterial genome data
  genome_taxonomy, 
  genome_stats, 
  genome_tree, 
  genome_kegg_paths,
  read_counts,
  
  # Transformed data
  mag_weighted,
  total,
  hel, 
  
  # Phyloseq objects
  phylo_seq,

  # Functions
  gifts_elements,
  gifts_functions,
  gifts_domains,
  
  # Colors
  taxonomy_colors,
  phylum_colors,
  order_colors,
  sampling_day_colors,
  
  # Define file path
  file = "data/data_mg.Rdata")
```

### Clean environment

```{r clean, comment = "", message = FALSE, warning = FALSE}
rm(list = ls())
```
