# Chapter 4 - MAG catalogue richness

## 4.1 - Load MG data

```{r load_data_mag_catalogue, message = FALSE, warning = FALSE, echo = FALSE}
load("data/data_mg.Rdata")
```

 
## 4.2 - Densities of overall relative abundances by phylum (Fig. 1a)

### Calculate densities

```{r calculate_density, message = FALSE, warning = FALSE, echo = FALSE}
sum_ind <-
  total %>%
  rownames_to_column(var = 'genome') %>%
  left_join(genome_taxonomy %>% select(genome, phylum), by = 'genome') %>%
  mutate(phylum = factor(phylum, levels = unique(names(phylum_colors)))) %>%
  pivot_longer(-c(genome, phylum), values_to = 'value', names_to = 'animal_code') %>%
  group_by(phylum, animal_code) %>%
  summarise(total_ind = sum(value), mean_ind = mean(value))

sum_phylum <-
  sum_ind %>%
  group_by(phylum) %>%
  summarise(total_phylum = sum(total_ind), mean_phylum = mean(mean_ind))
```

### Filter rare phylums 

```{r filter_phylums, warning = FALSE, comments = "", message = FALSE}
count_taxa <-
  sum_ind %>%
  filter(phylum %in% c('Bacillota_A',
                       'Bacillota',
                       'Bacteroidota',
                       'Cyanobacteriota',
                       'Pseudomonadota',
                       'Actinomycetota',
                       'Verrucomicrobiota',
                       'Campylobacterota'))
```

### Plot curves 

```{r density_histogram, comment = "", message = FALSE, warning = FALSE, fig.height = 8, fig.width = 7, fig.fullwidth = TRUE}
ggplot(count_taxa, aes(x = total_ind, colour = phylum, fill = phylum)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = phylum_colors) +
  scale_color_manual(values = phylum_colors) +
  scale_x_log10() +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = 'none')
```


## 4.3 - Functional landscape of the catalogue

### Caltulate t-SNE based on GIFTs

```{r tsne, warning = FALSE, comments = "", message = FALSE}
set.seed(100)
tsne_func <- Rtsne(X = gifts_elements, dims = 2, check_duplicates = FALSE)

genome_taxonomy <- 
  genome_taxonomy %>% 
  mutate(avg_mci = rowMeans(gifts_elements)) %>% 
  mutate(length = genome_stats$mag_length)

tsne_df <-
  tsne_func$Y %>%
  as.data.frame() %>%
  mutate(genome = rownames(gifts_elements)) %>%
  left_join(genome_taxonomy, by = 'genome') %>%
  mutate_at(vars(phylum, class, order, family), factor) %>%
  mutate(order = factor(order, levels = names(order_colors))) %>%
  mutate(phylum = factor(phylum, levels = names(phylum_colors))) %>%
  rename(tsne1 = 'V1', tsne2 = 'V2')

```

### Paint t-SNE by order and Overall Metabolic Capacity Index (MCI) (Figs. 1b and 1c)

```{r tsne_dotplot, comment = "", message = FALSE, warning = FALSE, fig.height = 4, fig.width = 12, fig.fullwidth = TRUE}
tsne_taxonomy <- 
  tsne_df %>%
  ggplot(aes(x = tsne1, y = tsne2, color = order)) +
  geom_point(size = 2, shape = 16, alpha = 0.8) +
  scale_color_manual(values = order_colors) +
  theme_minimal() +
  theme(legend.position = 'none')

tsne_mci <- 
  tsne_df %>%
  ggplot(aes(x = tsne1, y = tsne2, color = avg_mci)) +
  geom_point(size = 2, shape = 16, alpha = 0.8) +
  scale_colour_gradientn(colours = rev(terrain.colors(10))) +
  theme_minimal() +
  theme(legend.position = 'none')

grid.arrange(tsne_taxonomy, tsne_mci, nrow = 1)
```

### Interactive plot to go deeper into each MAG (not included in the paper)

```{r interactive_dotplot, comment = "", message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, fig.fullwidth = TRUE}
library(plotly)

p <-
  tsne_df %>%
  mutate(tooltip = paste(family, "//", genome)) %>% 
  ggplot(aes(x = tsne1, y = tsne2, colour = order, text = tooltip)) +
  geom_point(size = 4, stroke = 1, alpha = 0.8, shape = 16) + 
  scale_colour_manual(values = order_colors) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

### t-SNE coloring 4 most important phylums (Supplementary Fig. S1)

```{r tsne_by_phylum, comment = "", message = FALSE, warning = FALSE, fig.height = 8, fig.width = 12, fig.fullwidth = TRUE}
# Bacillota_A
tsne_a <- 
  tsne_df %>%
  ggplot(aes(x = tsne1, y = tsne2, colour = ifelse(phylum == 'Bacillota_A', as.character(order), 'other'))) +
  geom_point(size = 4, stroke = 1, alpha = 0.8, shape = 16) + 
  scale_colour_manual(values = c(
    Oscillospirales = "#fd8854",
    TANB77 = "#f4814d",
    Christensenellales = "#c26340",
    Lachnospirales = "#c28c5c",
    Clostridiales = "#ce9360",
    Monoglobales_A = "#ce734f",
    Peptostreptococcales = "#c65631",
    Monoglobales = "#b93725",
    UBA1212 = "#b10f19",
    other = "grey")) +
  theme_minimal() +
  labs(colour = "Bacillota_A") +
  theme(legend.position = 'bottom')

# Bacillota
tsne_b <- 
  tsne_df %>%
  ggplot(aes(x = tsne1, y = tsne2, colour = ifelse(phylum == 'Bacillota', as.character(order), 'other'))) +
  geom_point(size = 4, stroke = 1, alpha = 0.8, shape = 16) + 
  scale_colour_manual(values = c(
    RF39 = "#ecf76d",
    Lactobacillales = "#feef68",
    'ML615J-28' = "#fde671",
    Erysipelotrichales = "#ffd366",
    Acholeplasmatales = "#fdc151",
    Bacillales = "#fc953d",
    RFN20 = "#fd8035",
    other = "grey")) +
  theme_minimal() +
  labs(colour = "Bacillota") +
  theme(legend.position = 'bottom')

# Bacteroidota
tsne_c <- 
  tsne_df %>%
  ggplot(aes(x = tsne1, y = tsne2, colour = ifelse(phylum == 'Bacteroidota', as.character(order), 'other'))) +
  geom_point(size = 4, stroke = 1, alpha = 0.8, shape = 16) + 
  scale_colour_manual(values = c(
    Bacteroidales = "#4dc87c",
    Flavobacteriales = "#88c88b",
    'other' = "grey")) +
  theme_minimal() +
  labs(colour = "Bacteroidota") +
  theme(legend.position = 'bottom')

# Pseudomonadota
tsne_d <- 
  tsne_df %>%
  ggplot(aes(x = tsne1, y = tsne2, colour = ifelse(phylum == 'Pseudomonadota', as.character(order), 'other'))) +
  geom_point(size = 4, stroke = 1, alpha = 0.8, shape = 16) + 
  scale_colour_manual(values = c(
    Enterobacterales = "#60cfde",
    RF32 = "#60dfd2",
    Burkholderiales = "#5cdfb5",
    'Rs-D84' = "#40df91",
    'other' = "grey")) +
  theme_minimal() +
  labs(colour = "Pseudomonadota") +
  theme(legend.position = 'bottom')

grid.arrange(tsne_a, tsne_b, tsne_c, tsne_d, nrow = 2)
```


## 4.4 - Compare MCI between taxonomic groups (Supplementary Figs. S2 and S3)

```{r tidy_taxonomy_and_colors, comment = "", message = FALSE, warning = FALSE}
main_phylums <- c("Bacillota_A", "Bacillota", "Bacteroidota", "Cyanobacteriota","Pseudomonadota", "Actinomycetota", "Verrucomicrobiota", "Campylobacterota")

selected_taxonomy_colors <- 
  taxonomy_colors %>% 
  filter(phylum %in% main_phylums)

selected_orders <- 
  selected_taxonomy_colors %>% 
  select(order, color_order) %>% 
  filter(order != "Gastranaerophilales") %>% 
  filter(order != "Actinomycetales") %>% 
  filter(order != "Coriobacteriales")

selected_groups <- 
  selected_taxonomy_colors %>% 
  select(phylum, color_phylum) %>% 
  distinct(phylum, color_phylum) %>%
  filter(color_phylum != '#c28c5c') %>% 
  rename(order = 'phylum') %>% 
  rename(color_order = 'color_phylum') %>% 
  bind_rows(., selected_orders) %>% 
  mutate(order = factor(order, levels = c("Campylobacterota", 
                                          "Verrucomicrobiota",
                                          "Actinomycetota",
                                          "Enterobacterales",     
                                          "RF32",                 
                                          "Burkholderiales",      
                                          "Rs-D84",              
                                          "Pseudomonadota",  
                                          "Gastranaerophilales",
                                          "Cyanobacteriota", 
                                          "Bacteroidales",        
                                          "Flavobacteriales",
                                          "Bacteroidota", #
                                          "RF39",                 
                                          "Lactobacillales",      
                                          "ML615J-28",           
                                          "Erysipelotrichales",   
                                          "Acholeplasmatales",    
                                          "Bacillales",           
                                          "RFN20",
                                          "Bacillota",            
                                          "Oscillospirales",      
                                          "TANB77",               
                                          "Christensenellales",   
                                          "Lachnospirales",      
                                          "Clostridiales",        
                                          "Monoglobales_A",       
                                          "Peptostreptococcales", 
                                          "Monoglobales",        
                                          "UBA1212",
                                          "Bacillota_A"))) %>% 
  arrange(order)
```  

```{r tidy_data, comment = "", message = FALSE, warning = FALSE}
genome_taxonomy <- 
  genome_taxonomy %>% 
  mutate(avg_mci = rowMeans(gifts_elements)) %>% 
  mutate(length = genome_stats$mag_length)

table_one <- 
  genome_taxonomy %>% 
  select(genome, order, avg_mci)
  
table_two <- 
  genome_taxonomy %>% 
  select(genome, phylum, avg_mci) %>%
  rename(order = 'phylum') %>% 
  bind_rows(., table_one) %>% 
  rename(group = 'order')

border_groups <- c("Bacillota_A", "Bacillota", "Bacteroidota", "Cyanobacteriota",
                   "Pseudomonadota", "Actinomycetota", "Verrucomicrobiota", 
                   "Campylobacterota")
```
 
 ### Overall MCI by taxonomic phylum and order 
 
```{r mci_boxplot, comment = "", message = FALSE, warning = FALSE, fig.height = 7, fig.width = 6, fig.fullwidth = TRUE}
table_two %>% 
  filter(group %in% selected_groups$order) %>% 
  mutate(group = factor(group, levels = selected_groups$order)) %>% 
  ggplot() +
  geom_jitter(aes(x = group, y = avg_mci), color = 'grey', alpha = 0.5, width = 0.1) +
  geom_violin(aes(x = group, y = avg_mci, fill = group), color = "black", size = 0.01) +
  theme_minimal() +
  xlab("Taxonomic group") +
  ylab("Genome average metabolic capacity") +
  scale_fill_manual(values = selected_taxonomy_colors$color_order) +
  theme(legend.position = 'none',
        ) +
  coord_flip()
```  

### Genome length by taxonomic phylum and order 

```{r length_boxplot, comment = "", message = FALSE, warning = FALSE, fig.height = 7, fig.width = 6, fig.fullwidth = TRUE}
table_three <- 
  genome_taxonomy %>% 
  select(genome, order, length)

table_four <- 
  genome_taxonomy %>% 
  select(genome, phylum, length) %>%
  rename(order = 'phylum') %>% 
  bind_rows(., table_three) %>% 
  rename(group = 'order')

table_four %>% 
  filter(group %in% selected_groups$order) %>% 
  mutate(group = factor(group, levels = selected_groups$order)) %>% 
  ggplot() +
  geom_jitter(aes(x = group, y = length), color = 'grey', alpha = 0.5, width = 0.1) +
  geom_violin(aes(x = group, y = length, fill = group), color = "black", size = 0.01) +
  theme_minimal() +
  xlab("Taxonomic group") +
  ylab("Mag length") +
  scale_fill_manual(values = selected_taxonomy_colors$color_order) +
  theme(legend.position = 'none') +
  coord_flip()
```

### Clean environment

```{r clean_4, comment = "", message = FALSE, warning = FALSE}
rm(list = ls())
```
