# Chapter 8 - Temporal HMSC setup

HMSC models are computationally intensive. It is advisable to run this script on a server. The code is made to be reproduced on any supercomputer.

## 8.1 - Load MG data

### Define directories

```{r directories, message = FALSE, warning = FALSE, echo = FALSE}
# Create directories
dir.create("temporal_hmsc")
dir.create("temporal_hmsc/model_fit")
dir.create("temporal_hmsc/models")
dir.create("temporal_hmsc/panels")

# Define paths
localDir = "."
ModelDir = file.path(localDir, "temporal_hmsc/models")
PanelDir = file.path(localDir, "temporal_hmsc/panels")
```

### Generate input objects

```{r load_mg_data, message = FALSE, warning = FALSE, echo = FALSE}
# Load data
load("data/data_mg.Rdata")

# Metadata summary
design <- 
  chicken_metadata[, c('animal_code', 'trial', 'pen', 'age','breed',
                       'sex', 'treatment', 'chicken_body_weight')] %>% 
  column_to_rownames("animal_code")

design$log_seq_depth <- log(colSums(read_counts %>% column_to_rownames("genome")))
design$animal_code <- rownames(design)

for (i in 1:ncol(design)) {
  if (is.character(design[,i])) {
    design[,i] <- factor(design[,i])
  }
}
```


## 8.2- Prepate data for HMSC

### Define tables

```{r hmsc_tables, message = FALSE, warning = FALSE, echo = FALSE}
# Genome count table (quantitative community data)
YData <- log1p(mag_weighted %>% 
                 t() %>% 
                 as.data.frame())

# Fixed effects data (explanatory variables)
XData <- design
mean(rownames(YData) == rownames(XData))  # Ydata and XData in the same order

# Genome phylogeny
PData <- genome_tree

YData <- YData[,colnames(YData) %in% genome_tree$tip.label]  # Filter missing MAGs in tree
```

## Define formula

```{r hmsc_formumla, message = FALSE, warning = FALSE, echo = FALSE}
# Define XFormula
XFormula_Time <- ~log_seq_depth + trial + sex + breed + treatment + age

# Study Design
StudyDesign <- design[,c(2,9)]
rL.Pen <- HmscRandomLevel(units = levels(StudyDesign$pen))
```

### Define HMSC model

```{r hmsc_model, comment = "", echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
m <- Hmsc(Y = YData,
                XData = XData,
                XFormula = XFormula_Time,
                studyDesign = StudyDesign,
                ranLevels = list("pen" = rL.Pen),
                phyloTree = genome_tree,
                distr = "normal",
                YScale = TRUE)

save(m, file = file.path(ModelDir,"unfitted_model.Rdata"))
```

### Define MCMC characteristics

```{r mcmc_features, comment = "", echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
## How often to sample the MCM
samples_list = c(5, 250, 250)

# The number of MCMC steps between each recording sample
thin_list = c(1,1,10)
nst = length(thin_list)

# The number of MCMC chains ot use
nChains = 4

# Load only the unfitted models
load(file.path(ModelDir, "unfitted_model.Rdata"))
unf_model <- m
```


## 8.3 - Fit model

```{r fit_model, comment = "", echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
set.seed(1)

for (Lst in 1:length(samples_list)) {
  thin = thin_list[Lst]
  samples = samples_list[Lst]
  # Fit the model
  m = sampleMcmc(unf_model,
                 samples = samples,
                 thin = thin,
                 adaptNf = rep(ceiling(0.4*samples*thin), unf_model$nr),
                 transient = ceiling(0.5*samples*thin),
                 nChains = nChains,
                 nParallel = nChains)
  # Create file name
  filename = paste("temporal_model",
                   "_thin_", as.character(thin),
                   "_samples_", as.character(samples),
                   "_chains_", as.character(nChains),
                   ".Rdata", sep = "")
  # Save file
  save(m, file = file.path(ModelDir, filename)) 
}
```


## 8.4 - Evaluate convergence (Figure not added to Supplementary but could be included)

```{r convergence, comment = "", echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
set.seed(1)

for (i in 1) {
  ma = NULL
  na = NULL
  ma_rho = NULL
  na_rho = NULL
  for (Lst in 1:nst) {
    thin = thin_list[Lst]
    samples = samples_list[Lst]
    filename = paste("temporal_model","_thin_", as.character(thin),
                     "_samples_", as.character(samples),
                     "_chains_", as.character(nChains),
                     ".Rdata", sep = "")
    load(file = file.path(ModelDir, filename))
    mpost = convertToCodaObject(m,
                                spNamesNumbers = c(T,F),
                                covNamesNumbers = c(T,F))
    
    ## Beta - Species niches - response of MAGs to the fixed effects
    psrf.beta = gelman.diag(mpost$Beta,multivariate = FALSE)$psrf
    
    ## Rho - Influence of phylogeny on species niches - phylogenetic signal
    psrf.rho = gelman.diag(mpost$Rho,multivariate = FALSE)$psrf
    
    ## Beta
    if (is.null(ma)) {
      ma = psrf.beta[,1]
      na = paste0("temporal_model_", as.character(thin),
                  ",", as.character(samples))
    } else {
      ma = cbind(ma,psrf.beta[,1])
      na = c(na,paste0("temporal_model_", as.character(thin),
                       ",", as.character(samples)))
    }
    ## Rho
    if (is.null(ma_rho)) {
      ma_rho = psrf.rho[,1]
      na_rho = paste0("temporal_model_", as.character(thin),
                      ",", as.character(samples))
    } else {
      ma_rho = cbind(ma_rho,psrf.rho[,1])
      na_rho = c(na_rho,paste0("temporal_model_", as.character(thin),
                               ",", as.character(samples)))
    }
  }
  
  # Create file name
  panel.name = paste("MCMC_convergence",
                     "temporal_model_", as.character(i),
                     ".pdf", sep = "")
  # Plot
  pdf(file = file.path(PanelDir, panel.name))
  
  ## Beta
  par(mfrow = c(2,1))
  vioplot(ma,
          col = rainbow_hcl(1),
          names = na,
          ylim = c(0, max(ma)),
          main = "psrf(beta)")
  vioplot(ma,
          col = rainbow_hcl(1),
          names = na,
          ylim = c(0.9,1.1),
          main = "psrf(beta)")

  ## Rho
  par(mfrow = c(2,1))
  vioplot(ma_rho,
          col = rainbow_hcl(1),
          names = na_rho,
          ylim = c(0, max(ma_rho)),
          main = "psrf(rho)")
  vioplot(ma_rho,
          col = rainbow_hcl(1),
          names = na_rho,
          ylim = c(0.9,1.1),
          main = "psrf(rho)")
  dev.off()
}
```

### Clean environment

```{r clean, comment = "", message = FALSE, warning = FALSE}
rm(list = ls())
```
