# Chapter 11 - Associate microbial community metrics with chicken body weight

## 11.1 - Load MG and MT data

```{r load_data_mag_div_bw, message = FALSE, warning = FALSE, echo = FALSE}
load("data/data_mg.Rdata")
load("data/alpha_div.Rdata")
load("data/mci_com.Rdata")
```


## 11.2 - Associate alpha diversity metrics with chicken BW

### Neutral 

```{r plot_regression_n, warning =FALSE, comments = "", message = FALSE}
ggplot(alpha_div, aes(x = neutral, y = chicken_body_weight)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ sampling_time*trial, scales = 'free')
```

## Phylogenetic

```{r plot_regression_p, warning = FALSE, comments = "", message = FALSE}
ggplot(alpha_div, aes(x = phylogenetic, y = chicken_body_weight)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ sampling_time * trial, scales = 'free')
```

## Functional KEGG

```{r plot_regression_f, warning = FALSE, comments = "", message = FALSE}
ggplot(alpha_div, aes(x = functional_kegg, y = chicken_body_weight)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ sampling_time * trial, scales = 'free')
```


## 11.3 - Linear regressions 

### Neutral (Fig. 3a)

```{r calculate_lms_neutral, warning = FALSE, comments = "", message = FALSE}
div_all_day_35 <-
  alpha_div %>%
  filter(sampling_time == '35') %>%
  filter(trial != 'CC') %>% 
  filter(!animal_code %in% c("CB12.17", "CA04.16", "CA03.17")) # ¿? double check if we are OK filtering some samples, they show out of place diversity

N <- lme(chicken_body_weight ~ age + trial + neutral,
         random = ~1|pen,
         data = div_all_day_35)
summary(N)

plot_model(N,
           type = 'eff',
           title = 'Neutral diversity',
           terms = 'neutral',
           show.data = TRUE)
```

### Phylogenetic

```{r calculate_lms_p, warning = FALSE, comments = "", message = FALSE}
P <- lme(chicken_body_weight ~ age + trial + phylogenetic,
         random = ~1|pen,
         data = div_all_day_35)
summary(P)

plot_p <-
  plot_model(P,
             type = 'eff',
             title = 'Phylogenetic diversity',
             terms = 'phylogenetic',
             show.data = TRUE)
```

### Functional KEGG

```{r calculate_lms_f, warning = FALSE, comments = "", message = FALSE}
Q_kegg <- lme(chicken_body_weight ~ age + trial + functional_kegg,
         random = ~1|pen,
         data = div_all_day_35)
summary(Q_kegg)

plot_q <-
  plot_model(Q_kegg,
             type = 'eff',
             title = 'Functional diversity',
             terms = 'functional_kegg',
             show.data = TRUE)
```


## 11.4 - Associate community MCI with diversity

```{r mci_bw, warning = FALSE, comments = "", message = FALSE}
domains_com_2 <-
  funcs_com %>%
  as.data.frame(optional = TRUE) %>% 
  rownames_to_column(var = "animal_code") %>% 
  rowwise() %>%
  mutate(overall_com_mci = mean(c_across(B01:D09))) %>%
  select(animal_code, overall_com_mci) %>%
  left_join(alpha_div) %>%
  filter(sampling_time == '35')

# Neutral
N <- lme(neutral ~ age + trial + overall_com_mci, 
         random = ~1|pen,
         data = domains_com_2)
summary(N)

# Phylogenetic
P <- lme(phylogenetic ~ age + trial + overall_com_mci,
         random = ~1|pen,
         data = domains_com_2)
summary(P)

# Functional
Q <- lme(functional_kegg ~ age + trial + overall_com_mci,
         random = ~1|pen,
         data = domains_com_2)
summary(Q)
```


## 11.5 - Associate community MCI with chicken BW

```{r com_mci_vs_bw, warning = FALSE, comments = "", message = FALSE}
MCI <- lme(chicken_body_weight ~ age + trial + overall_com_mci,
         random = ~1|pen,
         data = domains_com_2)
summary(MCI)
```


# 11.6 - Associate community weighted genome size with chicken BW

```{r com_genome_size_vs_bw, warning = FALSE, comments = "", message = FALSE}
metadata_day_35 <-
  chicken_metadata %>%
  filter(sampling_time == "35")

mag_lengthed <- 
  round(sweep(total, MARGIN = 1, genome_stats$mag_length, `*`), 0) %>% 
  t() %>% 
  as.data.frame() %>%
  rownames_to_column(var = "animal_code") %>%
  rowwise() %>% 
  mutate(comm_length = mean(c_across(2:389))) %>% 
  select(animal_code, comm_length) %>% 
  filter(animal_code %in% metadata_day_35$animal_code) %>% 
  left_join(metadata_day_35, by = 'animal_code')

L <- lme(chicken_body_weight ~ scale(age) + trial + scale(log(comm_length)),
         random = ~ 1|pen,
         data = mag_lengthed)
summary(L)
plot(L)
```

### Clean environment

```{r clean_11, comment = "", message = FALSE, warning = FALSE}
rm(list = ls())
```
