# Chapter 13 - Bacteria vs Chicken BW HMSC analysis

## 13.1 - Load model

### Define directories

```{r directories, comment = "", echo = FALSE}
# Directory
localDir = "."
ModelDir = file.path(localDir, "hmsc_bw/models")
PanelDir = file.path(localDir, "hmsc_bw/panels")
MFDir <- file.path(localDir, "hmsc_bw/model_fit")
```

### Load model
```{r load_model, comment = "", echo = FALSE}
load(file = file.path(ModelDir,"model_thin_10_samples_250_chains_4.Rdata"))
```


## 13.2 - Evaluate model

```{r evalute_model, echo=FALSE}
partition <- createPartition(m, nfolds = 4)
preds <- computePredictedValues(m, partition = partition, nChains = 1) # nParallel = nChains
MFCV <- evaluateModelFit(hM = m, predY = preds)

save(MFCV,file = file.path(MFDir, "MF_thin_10_samples_250_chains_4.Rdata"))
```


## 13.3 - Examine species responses (Plot not included in Supplementary)

```{r mag_response, echo = FALSE}
beta_post <- getPostEstimate(m, parName = "Beta")

plotBeta(m, beta_post,supportLevel = 0.9, plotTree = TRUE)
```


## 13.4 - Examine variance partition (Plot not included in Supplementary)

```{r mag_response, echo = FALSE}
varpartition <- computeVariancePartitioning(m)

plotVariancePartitioning(m, VP = varpartition)

bw_var <- varpartition$vals[3,]
bw_param <- beta_post$mean[4,]
bw_support <- beta_post$support[4,]
```

## Create a data table

```{r mag_response, echo = FALSE}
MFCV$R2

rel_abu <- colMeans(vegan::decostand(exp(m$Y), method = "total")) * 100

toplot <- data.frame(parameter = bw_param,
                     bw_support = bw_support,
                     var = bw_var,
                     pred = MFCV$R2,
                     var_pred = bw_var * MFCV$R2)

toplot$MAG <- rownames(toplot)
toplot$rel_abu <- rel_abu
rownames(toplot) <- NULL

write.table(toplot, file = "data/data.txt", row.names = F)
```

### Associate bacteria response with chicken BW (not included in Supplementary)

```{r mag_response_chicken_bw, echo = FALSE}
ggplot() +
  geom_point(data = toplot, aes(x = parameter, y = Pred), alpha = 0.7, shape = 16) +
  geom_point(data = toplot[toplot$bw_support > 0.975,],
             mapping = aes(x = parameter, y = Pred),
             color = "red", shape = 16) +
  geom_point(data = toplot[toplot$bw_support < 0.025,],
             mapping = aes(x = parameter, y = Pred),
             color = "blue", shape = 16) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = mean(toplot$Pred, na.rm = TRUE), linetype = "dashed") +
  labs(x = "Parameter estimate", y = "Body weight's importance") +
  theme_minimal()
```


## 13.5 - Create table with HMSC results

```{r create_table, echo = FALSE}
# Total relative abundance of species associated positively and negatively with body weight (~9%)
sum(toplot$rel_abu[(toplot$bw_support > 0.975 | toplot$bw_support < 0.025) & (toplot$var_pred > mean(toplot$var_pred))])

# Total relative abundance of species associated positively with body weight (<1%)
sum(toplot$rel_abu[(toplot$bw_support > 0.975) & (toplot$var_pred > mean(toplot$var_pred))])

# Total relative abundance of species associated negatively with body weight (>8%)
sum(toplot$rel_abu[(toplot$bw_support < 0.025 | toplot$bw_support < 0.025) & (toplot$var_pred > mean(toplot$_pred))])

# Phylogenetic signal
mpost <- convertToCodaObject(m)
quantile(unlist(mpost$Rho), probs = c(.05,.5,.95))

bw_parameter <- beta_post$mean[4,]

bw_parameter_phyloSorted <-
  data.frame(bw_parameter = bw_parameter[
    match(m$phyloTree$tip.label,
          names(bw_parameter))])

mean(rownames(bw_parameter_phyloSorted) == m$phyloTree$tip.label)

new_tree <- m$phyloTree
new_tree$node.label <- NULL

obj <- phylo4d(new_tree, tip.data = bw_parameter_phyloSorted)
barplot.phylo4d(obj)

bw.cg <- phyloCorrelogram(obj, trait = "bw_parameter")
saveRDS(bw.cg,file = file.path(PanelDir,"bw.cg.rds"))
bw.cg <- readRDS(file = file.path(PanelDir,"bw.cg.rds"))

plot(bw.cg)

bw_parameter_df <- data.frame(genome = names(bw_parameter), bw_response = bw_parameter)
bw_parameter_df$bw_trend <- NA
bw_parameter_df$bw_trend[beta_post$support[4,] > 0.95] <- "positive"
bw_parameter_df$bw_trend[beta_post$support[4,] < 0.05] <- "negative"
bw_parameter_df$bw_trend[is.na(bw_parameter_df$bw_trend)] <- "neutral"

write_tsv(bw_parameter_df,file = "data/hmsc_bw_trend.tsv")
```

### Clean environment

```{r clean, comment = "", message = FALSE, warning = FALSE}
rm(list = ls())
```
