# Testing completeness correction before calculating distilling GIFTs

## Load MG data

```{r load_mg_data_supp, warning = FALSE, comments = "", message = FALSE}
load("data/data_mg.Rdata")

# Load raw genome annotations
genome_annotations <- read_tsv("data/gene_annotations.tsv.xz")

gifts_raw <- distill(genome_annotations, GIFT_db2,
                     genomecol = 2, annotcol = c(9,10,19), verbosity = F)

# Load ENA to genome_id table
ena_to_mag_id <- read_tsv("data/ena_to_mag_id.tsv")

gifts_matrix <-
  gifts_raw %>%
  data.frame() %>%
  rownames_to_column(var = 'mag_name') %>%
  left_join(ena_to_mag_id %>%
              select(mag_name, mag_id),
            by = 'mag_name') %>%
  select(-mag_name) %>%
  relocate(mag_id) %>%
  filter(mag_id %in% genome_taxonomy$genome) %>%
  column_to_rownames('mag_id') %>%
  as.matrix()

genome_completeness <-
  genome_stats %>%
  select(genome, completeness_score) %>%
  as.matrix()
```

## Define function for correction

```{r define_function, warning = FALSE, comments = "", message = FALSE}
# Correction function defined in
# https://github.com/anttonalberdi/DAMMA/blob/main/R/damma_correction.R
# Based on https://www.nature.com/articles/s43705-023-00221-z
gifts_correction <- function(MCI_table, genome_completeness, stats = TRUE) {
  #Sort Genomes and test matching
  MCI_table <- MCI_table[order(rownames(MCI_table)),]
  genome_completeness <- genome_completeness[order(genome_completeness[,1]),]
  if (all(as.character(rownames(MCI_table)) !=
          as.character(genome_completeness[,1])))
    stop("Pathway table is missing")
  #Get completeness values
  Genome_completeness <- as.numeric(genome_completeness[,2])
  #Create corrected fullness matrix
  MCI_table_corrected <- matrix(0,nrow = nrow(MCI_table),ncol = ncol(MCI_table))
  colnames(MCI_table_corrected) = colnames(MCI_table)
  rownames(MCI_table_corrected) = rownames(MCI_table)
  #Iterate modelling and correction for each function
  suppressWarnings(
    for (i in 1:ncol(MCI_table)) {
      Model <- glm(MCI_table[,i]~Genome_completeness,family = "binomial")
      slope_coef <- coef(Model)[2]
      if (slope_coef > 0) {
        for (j in 1:nrow(MCI_table)) {
          # Model prediction of fullness if completeness was 100%
          pred_100 <- round(predict(Model,
                                    newdata = data.frame(
                                      Genome_completeness = 100),
                                    type = "response"),1)
          # Model prediction of fullness for actual completeness of the focal MAG
          pred_focal <- round(predict(Model,
                                      newdata = data.frame(
                                        Genome_completeness = Genome_completeness[j]),
                                      type = "response"), 1)
          # The expected change in function fullness if focal MAG was 100% complete
          pred_diff <- pred_100 - pred_focal
          MCI_table_corrected[j,i] = MCI_table[j,i] + pred_diff
        }
      } else if (slope_coef <= 0) {
        MCI_table_corrected[,i] <- MCI_table[,i]
      }
    }
  )
  # If corrected fullness >1, convert it to 1.
  MCI_table_corrected[MCI_table_corrected > 1] <- 1
  # Outout overall correction statistics on screen
  total <- nrow(MCI_table)*ncol(MCI_table)
  changes <- c(MCI_table == MCI_table_corrected)
  changes <- length(changes[!(changes)])
  percentage <- round(changes / total * 100,1)
  cat(paste0(changes," out of ",total," (",percentage,"%) fullness values were corrected\n"))
  if (stats == TRUE) {
    #Outout overall correction statistics on screen
    total <- ncol(MCI_table_corrected)
    for (r in rownames(MCI_table_corrected)) {
      completeness <- round(as.numeric(genome_completeness[genome_completeness[,1] == r,2]),1)
      changes <- MCI_table[r,] == MCI_table_corrected[r,]
      changes <- length(changes[!(changes)])
      percentage <- round(changes / total * 100,1)
      cat(paste0("\t",r," (",completeness,"%): ",changes,
                 "/",total," (",percentage,"%) fullness values were corrected\n"))
    }
  }
  #Output corrected table
  return(MCI_table_corrected)
}
```

## Calculate corrected GIFTs

```{r correct_gifts, warning = FALSE, comments = "", message = FALSE}
gifts_corrected <- gifts_correction(gifts_matrix, genome_completeness)

# calculate GIFTs by element, function and domain
gifts_elements_corrected <- to.elements(gifts_corrected, GIFT_db2)

gifts_functions_corrected <- to.functions(gifts_elements_corrected, GIFT_db2)

gifts_domains_corrected <- to.domains(gifts_functions_corrected, GIFT_db2)
```

## Save files so you don't spend time re-generating them
```{r save_gift_corr_files, warning = FALSE, comments = "", message = FALSE}
save(
  gifts_raw,
  gifts_corrected,
  gifts_elements_corrected,
  gifts_functions_corrected,
  gifts_domains_corrected,
  file = "data/correction_files.Rdata")
```

## compare MCI values without and with correction

```{r ave_mci_by_mag, warning = FALSE, comments = "", message = FALSE}
# Load data

load("data/correction_files.Rdata")

# calculate average MCI by MAG

# a) without correction
overall_mci <-
  gifts_elements %>%
  as.data.frame() %>%
  rownames_to_column(var = "genome") %>%
  rowwise() %>%
  mutate(mean_MCI = mean(c_across(B0101:S0301))) %>%
  select(genome, mean_MCI) %>%
  left_join(genome_stats, by = "genome")%>%
  mutate(mag_length = mag_length / 1000000) %>%
  left_join(genome_taxonomy, by = "genome")

# b) Applying correction
overall_mci_corrected <-
  gifts_elements_corrected %>%
  as.data.frame() %>%
  rownames_to_column(var = "genome") %>%
  rowwise() %>%
  mutate(mean_MCI = mean(c_across(B0101:S0301))) %>%
  select(genome, mean_MCI) %>%
  left_join(genome_stats, by = "genome") %>%
  mutate(mag_length = mag_length / 1000000) %>%
  left_join(genome_taxonomy, by = "genome")
```

## Spearman correlation

```{r corr_mci_1, warning = FALSE, comments = "", message = FALSE}
# a) without correction
cor(overall_mci$mean_MCI, overall_mci$completeness_score, method = "spearman")
```

```{r corr__mci_2, warning = FALSE, comments = "", message = FALSE}
# b) Applying correction
cor(overall_mci_corrected$mean_MCI, overall_mci_corrected$completeness_score, method = "spearman")
```

## Comparing average MCI with genome completeness

```{r ploting_correc_1, warning = FALSE, comments = "", message = FALSE, fig.height = 4, fig.width = 6, fig.fullwidth = TRUE}
# a) Without correction
overall_mci %>%
  ggplot(aes(x = completeness_score, y = mean_MCI, color = order)) +
  geom_point(size = 2.5) +
  xlim(50,100) + ylim(0, 0.6) +
  scale_color_manual(values = order_colors) +
  theme_bw() +
  theme(legend.position = 'none') +
  xlab("Completeness score") + ylab("Mean MCI") +
  ggtitle("Without correction")
```

```{r ploting_correc_2, warning = FALSE, comments = "", message = FALSE, fig.height = 4, fig.width = 6, fig.fullwidth = TRUE}
# b) Applying correction
overall_mci_corrected %>%
  ggplot(aes(x = completeness_score, y = mean_MCI, color = order)) +
  geom_point(size = 2.5) +
  xlim(50,100) + ylim(0, 0.6) +
  scale_color_manual(values = order_colors) +
  xlab("Completeness score") + ylab("Mean MCI") +
  theme_bw() +
  theme(legend.position = 'none') +
  ggtitle("Applying correction")
```

## Reduced genomes

```{r reduced_genomes_only, warning = FALSE, comments = "", message = FALSE}
# a) Without correction
reduced_genomes <-
  overall_mci %>%
  filter(order == "RF39" | family == "UBA1242")

# b) Applying correction
reduced_genomes_corrected <-
  overall_mci_corrected %>%
  filter(order == "RF39" | family == "UBA1242")
```

## Spearman correlation

```{r cor_red_1, warning = FALSE, comments = "", message = FALSE}
# a) Without correction
cor(reduced_genomes$mean_MCI, reduced_genomes$completeness_score, method = "spearman")
```

```{r cor_red_2, warning = FALSE, comments = "", message = FALSE}
# b) Applying correction
cor(reduced_genomes_corrected$mean_MCI, reduced_genomes_corrected$completeness_score, method = "spearman")
```

## Illustrating correlation between average MCI and completeness

```{r same_plot_but_red_genomes, warning = FALSE, comments = "", message = FALSE, fig.height = 4, fig.width = 6, fig.fullwidth = TRUE}
# a) Without correction
reduced_genomes %>% 
  ggplot(aes(x = completeness_score, y = mean_MCI, color = order)) +
  geom_point(size = 2.5) +
  xlim(50,100) + ylim(0, 0.6) +
  scale_color_manual(values = order_colors) +
  theme_bw() +
  theme(legend.position = 'none') +
  xlab("Completeness score") + ylab("Mean MCI") +
  ggtitle("Without correction")
```

```{r same_plot_but_red_genomes_2, warning = FALSE, comments = "", message = FALSE, fig.height = 4, fig.width = 9, fig.fullwidth = TRUE}
# b) Applying correction
reduced_genomes_corrected %>% 
  ggplot(aes(x = completeness_score, y = mean_MCI, color = order)) +
  geom_point(size = 2.5) +
  xlim(50,100) + ylim(0, 0.6) +
  scale_color_manual(values = order_colors) +
  xlab("Completeness score") + ylab("Mean MCI") +
  theme_bw() +
  theme(legend.position = 'none') +
  ggtitle("Applying correction")
```

## Testing differences at community-weighted MCI

```{r ploting_com_1, warning = FALSE, comments = "", message = FALSE, fig.height = 4, fig.width = 6, fig.fullwidth = TRUE}
# Load community-weighted MCI without correction
load("data/mci_com.Rdata")

funcs_com %>%
  as.data.frame() %>%
  rownames_to_column('animal_code') %>%
  rowwise() %>%
  mutate(overall = mean(c_across(B01:D09))) %>%
  select(animal_code, overall) %>%
  left_join(chicken_metadata %>%
              select(animal_code, sampling_time),
            by = 'animal_code') %>%
  ggplot(aes(x = sampling_time,
             y = overall,
             group = sampling_time)) +
  geom_jitter(aes_string(colour = 'sampling_time'), size = 0.3, width = 0.2, alpha = 0.6) +
  geom_boxplot(aes_string(fill = 'sampling_time'), width = 0.4, alpha = 0.5, outlier.color = NA) +
  scale_fill_manual(values = c('#E69F00', '#CC6677', '#56B4E9')) +
  scale_color_manual(values = c('#E69F00', '#CC6677', '#56B4E9')) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        legend.position = 'none') +
  ggtitle("Without correction")
```

```{r ploting_com_2, warning = FALSE, comments = "", message = FALSE, fig.height = 4, fig.width = 9, fig.fullwidth = TRUE}
# Calculate community-weighted MCI applying correction
funcs_com_corrected <- to.community(gifts_functions_corrected, total, GIFT_db2)

funcs_com_corrected %>%
  as.data.frame() %>%
  rownames_to_column('animal_code') %>%
  rowwise() %>%
  mutate(overall = mean(c_across(B01:D09))) %>%
  select(animal_code, overall) %>%
  left_join(chicken_metadata %>%
              select(animal_code, sampling_time),
            by = 'animal_code') %>%
  ggplot(aes(x = sampling_time,
             y = overall,
             group = sampling_time)) +
  geom_jitter(aes_string(colour = 'sampling_time'), size = 0.3, width = 0.2, alpha = 0.6) +
  geom_boxplot(aes_string(fill = 'sampling_time'), width = 0.4, alpha = 0.5, outlier.color = NA) +
  scale_fill_manual(values = c('#E69F00', '#CC6677', '#56B4E9')) +
  scale_color_manual(values = c('#E69F00', '#CC6677', '#56B4E9')) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        legend.position = 'none') +
  ggtitle("Applying correction")
```

```{r clean_16, comment = "", message = FALSE, warning = FALSE}
rm(list = ls())
```
